Predicting the PCG Pseudo-Random Number Generator In Practice

### 2020-08-19 article/main.tex
remarques

% !TeX spellcheck = en
\documentclass[preprint,svgnames,journal=tosc]{iacrtrans}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites; an initial
battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
edition of \emph{The Art of Computer Programming}. In 1996, Knuth's tests were then
supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
the \texttt{TestU01}~\cite{LEcuyerS07} library, whose ``BigCrush'' test is
considered state-of-the-art by the relevant community, to the best of our
knowledge. In 2010, the NIST proposed its own statistical test suite (Special
Publication 800-22), to which improvements were later
suggested~\cite{ZhuMLZJ16}. We understand that the \texttt{PractRand} test suite
also has a good reputation.

@@

Our algorithm reconstruct the internal state using a ``guess-and-determine''
approach: some bits of the internal state are guessed ; assuming the guesses are
correct, some other information is computed ; a consistency check discards bad
guesses early on ; then candidate internal states are computed and fully
tested. The problem actually come in two distinct flavors.

@@

Let $U$ be a vector of $n$ integers modulo $2^k$ such that
$U_{i+1} = a U_i \bmod 2^k$; it is a geometric progression of common ratio $a$,
modulo $2^k$, therefore $(U_0, \dots, U_{n-1}) \in \mathcal{L}$. Let
$T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$, and let $N$ denote an
arbitrary ``noise vector'' such that $N_i \in \{-1, 0, 1\}$. Finally, set
$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.  We will be facing the following
problem (``reconstructing noisy truncated geometric series'') several times:
\begin{description}
\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^\ell}\right)^n$, a (noisy) version of $U$ truncated to the top $\ell$ bits.
\item[OUTPUT] $U_0 \bmod 2^k$, the first term of the (non-truncated) geometric sequence.
\end{description}

We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
and many ``low-dimension'' in sections~\ref{sec:Cknown}
and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to
solve these problems. We first claim that $2^{k-\ell} \widetilde{T}$ is
``close'' to a point of the lattice $\mathcal{L}_{n,k}$.

\begin{lemma}\label{thm:close}
  There exists $U' \in \mathcal{L}$ such that $U'_i \equiv U_i \bmod 2^k$
  and $\left\lVert U' -2^{k-\ell} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-\ell}$.
\end{lemma}

\begin{proof}
  We first observe that $U$ belongs to the lattice $\mathcal{L}$. We start by setting $U' \gets U$, and we examine all coordinates of $U$:
  \begin{itemize}
  \item If $\widetilde{T}_i = T_i + N_i$ (without modulo), then we have:
    \begin{align*}
      \left| 2^{k-\ell}\widetilde{T}_i-U_i \right| &= \left| 2^{k-\ell}(T_i + N_i)-U_i \right| 
                                                     = \left| 2^{k-\ell}N_i - U_i[0,k-\ell] \right| \\
                                                   &\leq \left| 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \right| 
                                                   % &\leq 2^{k-\ell} + 2^{k-\ell} - 1\\
      \leq 2^{k-\ell +1 }.
    \end{align*}

  \item Otherwise, there are two possible ``wraparound'' cases:
    \begin{itemize}
    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^\ell-1$. 
      In this case, we have $U_i = U_i[0,k-\ell]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
      \[
        \left| 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \right| =
        \left| 2^{k-\ell} + U_i[0,k-\ell]\right| \leq 2^{k-\ell + 1}. 
      \]

    \item Or $T_i = 2^{\ell}-1$ and $N_i = +1$, which leads to
      $\widetilde{T}_i = 0$.  This implies that
      $U_i= 2^k-2^{k-\ell} + U_i[0,k-\ell]$; we set $U'_i = U_i - 2^k$ (again,
      with this modification $U'$ stays in the lattice), and we find:
      \[
        \left| 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
        = \left| 2^k-2^{k-\ell} + U_i[0,k-\ell] - 2^k\right| \leq 2^{k-\ell + 1}.
      \]
    \end{itemize}
  \end{itemize}
  In the end, we have
  $\left\| U' -2^{k-\ell} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-\ell}$,
  and $U' \equiv U \mod 2^k $.
\end{proof}

@@

Let $\CVP(\mathcal{L}, x)$ denote the vector of $\mathcal{L}$ closest to the
input vector $x$. Using the same notations as above, we want to know if
$\CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T}\right)$ is indeed $U'$. This
will necessarily be the case when $\|2^{k-\ell}\widetilde{T} - U'\|$ is smaller
than the length of the shortest non-zero vector of $\mathcal{L}$ --- this quantity,
the \emph{first minimum} of the lattice, is denoted by $\lambda_1(\mathcal{L})$.
%What can we say about $|CVP(L,2^{k-\ell}\widetilde{T})-U'|$ ?
By the triangular inequality, we have:
\begin{align*}
|\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T} \right) - 2^{k-\ell}\widetilde{T}\right | + \left|2^{k-\ell}\widetilde{T} - U'\right|.
\end{align*}
But, as the vector $U'$ belongs to the lattice, by definition of the closest
vector and by lemma~\ref{thm:close}:
% $|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leq
% |2^{k-\ell}\widetilde{T}-U'| $, hence
\[
  |\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
\]

If we can prove that the right side of this inequality is smaller than the first
minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
$\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T})$ indeed reveals $U_0 \bmod 2^k$.

In section~\ref{sec:big_cvp}, we will be facing the problem of reconstructing a
geometric sequence modulo $2^{128}$ given arbitrarily many (noisy versions of
the) most-significant $6$ bits of successive elements of the sequence. Therefore
we have $k=128$ and $\ell=6$, and we wish to determine the required number of
samples, \textit{i.e.} the value of $n$. This means finding the values of $n$
such that $\sqrt{n} 2^{124} \leq \lambda_1\left(\mathcal{L}\right)$.

Starting from $n=\lceil 122/6\rceil\rceil$, we computed the length of the
shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
until the condition holds. The \textsc{Shortest Vector Problem} (SVP) is another
well-known lattice NP-hard problem; we used the (almost) off-the-shelf
\texttt{G6K} library~\cite{AlbrechtDHKPS19}, which gave results very quickly by
sieving. \texttt{fplll}~\cite{fplll} was too slow above dimension 50, in the
default settings.

After this computation, we found that the minimal possible $n$ is 63: with
$n=63$, the shortest vector of $\mathcal{L}$ has length greater than
$2^{127.02}$, which is high enough. This vector can be obtained by bootstrapping
the geometric sequence with
\[
  U_0 = 12144252875850345479015002205241987363
\]
then reducing the terms modulo $2^{128}$ in zero-centered representation
(subtracting $2^{128}$ to $U_i$ if $U_i > 2^{127}$). It follows that when
$n\geq 63, k=128$ and $\ell=6$, any CVP oracle will return a vector congruent to
the original $U$ when given $\widetilde{T}$.

@@

Denote again by $\mathcal{L}$ the $n$-dimensional lattice spanned by the rows of
$G_{n,64}$, and let $H$ denote the LLL-reduction of $G_{n,64}$. The same lattice
is also spanned by the rows of $H$. For instance, with $n=3$:
\[
  H = \left(\begin{array}{rrr}
                -1241281756092 & 3827459685972 & -728312298332 \\
                -5001120657083 & -2117155768935 & 5479732607037 \\
                8655886039732 & 3303731088004 & 6319848582548
              \end{array}\right)
\]
Set $S = 2^{k-\ell} \widetilde{T}H^{-1}$; as $2^{k-\ell}\widetilde{T}$ is not
\textit{a priori} an element of the lattice, $S$ need not be an integer
vector. Let then $R$ denote the rounding of $S$, \textit{i.e.}
$R_i = \left\lfloor S_i \right\rceil$. Then $R H$ is an element of the
lattice. Under the right conditions, it will be the vector of the lattice
closest to $2^{k-\ell} \widetilde{T}$. Indeed:
\begin{align*}
\lVert U' - R H \rVert &= \left\| U' - \left(R - 2^{k-\ell}\widetilde{T}H^{-1} + 2^{k-\ell}\widetilde{T}H^{-1}\right)H \right\| \\
&= \left\| U' - 2^{k-\ell}\widetilde{T} - \left(R - 2^{k-\ell} \widetilde{T}H^{-1}\right )H \right\| \\
&\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\| R - 2^{k-\ell}\widetilde{T}H^{-1} \right\| \times \vvvert H \vvvert.
\end{align*}
By definition, $R$ is the closest integer vector to
$2^{k-\ell}\widetilde{T}H^{-1}$. But as $U'$ is an element of the lattice,
$U'H^{-1}$ is an integer vector. Thus $R-2^{k-\ell}\widetilde{T}G^{-1}$ is
shorter than $U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}$. Hence :
\begin{align*}
  \left\| U' - RH \right\| &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|R - 2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U'H^{-1}-2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U' - 2^{k-\ell}\widetilde{T}\right\| \times \left\vvvert H^{-1} \right\vvvert  \times \vvvert H \vvvert \\
                        & 	\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| \times \left(1 + \vvvert H^{-1} \vvvert  \times \vvvert H \vvvert \right).
\end{align*}

Note that $\vvvert H^{-1} \vvvert \times \vvvert H \vvvert$ is the condition
number of the matrix $H$. Lattice reduction has the side effect of reducing the
condition number, therefore it makes sense to use an LLL-reduced basis. If we
can prove that the right side of the inequality is smaller than the first
minimum of the lattice, then we would have proved that $RH$ is indeed the
closest vector we were searching for. Because we have fixed $k = 64$, by
lemma~\ref{thm:close} we have
$\left\| 2^{k-\ell}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-\ell}-1
$. So, if we fix $n$ we can search for the minimum number $\ell$ of known
most-significant bits such that:
\[
  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-\ell} \leq \lambda_1(\mathcal{L})
\]

\begin{table}[htbp!]
  \centering
  \begin{tabular}{|c|Sc|Sc|c|}
	\hline
	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $\ell$ \\
	\hline
	3 & 2.87 & $4.09e^{12} \simeq 2^{41.9}$ & 26 \\
	4 & 2.06 & $2.44e^{14} \simeq 2^{47.8}$ & 20 \\
	5 & 3.77 & $1.72e^{15}\simeq 2^{50.6}$ & 18 \\
	6 & 2.69 & $1.03e^{16}\simeq 2^{53.2}$ & 15 \\
	\hline
  \end{tabular}
  \caption{minimal $\ell$ needed for a given $n$}
  \label{tab:babai}
\end{table}

When $\ell$ is greater than the values given in table~\ref{tab:babai}, then
Babai's rounding technique will always return the closest vector, and allow us
to reconstruct a truncated geometric series.

@@

Adapting the previous reasoning enables an efficient state reconstruction
algorithm for the \texttt{lehmer64()} generator shown in
Fig.~\ref{fig:prngs}. When clocked, it outputs the top 64 bits of a geometric
sequence ($k=128$ and $\ell=64$). Three successive outputs are sufficient to
reconstruct the internal state using Babai's rounding technique. This yields the
following reconstruction algorithm:
\begin{minted}[fontsize=\scriptsize]{python}
def reconstruct(X):
    """
    Produce the internal state of the generator given three consecutive outputs of lehmer64().
    16 multiplications, 1 division, 11 additions and 3 roundings only.
    """
    a = 0xda942042e4dd58b5
    r = round(2.64929081169728e-7 * X[0] + 3.51729342107376e-7 * X[1] + 3.89110109147656e-8 * X[2])
    s = round(3.12752538137199e-7 * X[0] - 1.00664345453760e-7 * X[1] - 2.16685184476959e-7 * X[2])
    t = round(3.54263598631140e-8 * X[0] - 2.05535734808162e-7 * X[1] + 2.73269247090513e-7 * X[2])
    u = r * 1556524 + s * 2249380 + t * 1561981
    v = r * 8429177212358078682 + s * 4111469003616164778 + t * 3562247178301810180
    state = (a*u + v) % (2**128)
    return state
\end{minted}

@@

This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
dozen arithmetic operations; it always output the correct value of $S_0$, and
may output a few other ones (they can be easily discarded by checking $X_2$). An
improved ``guess-and-determine'' state reconstruction algorithm is possible,
which essentially amounts to expose a truncated version of the underlying linear
congruential generator, and attack it using the tools exposed in
section~\ref{sec:geometric}. This is possible by combining the following
ingredients:
\begin{itemize}
\item The underlying linear congruential generator uses a power-of-two modulus,
  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
  the $\ell$ low-order bits of $S_i$. More precisely, we have:
  \begin{equation}\label{eq:lcg}
    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
  \end{equation}
  Therefore, guessing the least-significant bits of $S_0$ yields a ``long-term
  advantage'' that holds for all subsequent states.

@@

We consider the sequence of internal states
$S = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
least-significant bits of $S_0$, therefore let us assume that their value is
known and denote it by $w$. We define $S' = \LCG_{128}(S_0 - w, 0)$ and
$K = \LCG_{128}(w, c)$ --- this is Known. By~\eqref{eq:lcg-additive}, we have
$S' = S - K$. The point is that the elements of $S'$ follow a geometric
progression of common ratio $a$; in addition, the $\ell$ least significant bits
of each components are equal to zero. It follows that $S'[\ell:64+\ell]$ also
follows a geometric progression of common ratio $a$, this time modulo
$2^{64}$. The crux of the reconstruction algorithm is to find
$S'[\ell:64+\ell]$.

We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
unknown vector $B$ of \emph{borrows}, whose components are either 0 or 1, such
that $T' = S[58:64+\ell] \boxminus K[58:64+\ell] \boxminus B$. Because the
borrows are unknown, we in fact compute
$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, and clearly
$\widetilde{T'} = T' \boxplus B$. We are thus in the context of the problem
discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
unknown vector $B$ of borrows. %Notice that the noise is asymmetric.

>>>>

% !TeX spellcheck = en
\documentclass[submission,svgnames,journal=tosc]{iacrtrans}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' if it passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites; an initial
battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
edition of \emph{The Art of Computer Programming}. In 1996, Knuth's tests were then
supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
the \texttt{TestU01}~\cite{LEcuyerS07} library, whose ``BigCrush'' test is
considered state-of-the-art by the relevant community, to the best of our
knowledge. In 2010, the NIST proposed its own statistical test suite (Special
Publication 800-22), to which improvements were later
suggested~\cite{ZhuMLZJ16}. We understand that the \texttt{PractRand} test suite
also has a good reputation.

@@

Our algorithm reconstructs the internal state using a ``guess-and-determine''
approach: some bits of the internal state are guessed ; assuming the guesses are
correct, some other information is computed ; a consistency check discards bad
guesses early on ; then candidate internal states are computed and fully
tested. The problem actually comes in two distinct flavors.

@@

Let $U$ be a vector of $n$ integers modulo $2^k$ such that
$U_{i+1} = a U_i \bmod 2^k$; it is a geometric progression of common ratio $a$,
modulo $2^k$, therefore $(U_0, \dots, U_{n-1}) \in \mathcal{L}$. Let
$T_i = U_i[k-t:k]$ denote the top $t$ bits of $U_i$, and let $N$ denote an
arbitrary ``noise vector'' such that $N_i \in \{-1, 0, 1\}$. Finally, set
$\widetilde{T}_i = T_i + N_i \bmod 2^{t}$.  We will be facing the following
problem (``reconstructing noisy truncated geometric series'') several times:
\begin{description}
\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^t}\right)^n$, a (noisy) version of $U$ truncated to the top $t$ bits.
\item[OUTPUT] $U_0 \bmod 2^k$, the first term of the (non-truncated) geometric sequence.
\end{description}

We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
and many ``low-dimension'' in sections~\ref{sec:Cknown}
and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to
solve these problems. We first claim that $2^{k-t} \widetilde{T}$ is
``close'' to a point of the lattice $\mathcal{L}$.

\begin{lemma}\label{thm:close}
  There exists $U' \in \mathcal{L}$ such that $U'_i \equiv U_i \bmod 2^k$
  and $\left\lVert U' -2^{k-t} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-t}$.
\end{lemma}

\begin{proof}
  We first observe that $U$ belongs to the lattice $\mathcal{L}$. We start by setting $U' \gets U$, and we examine all coordinates of $U$:
  \begin{itemize}
  \item If $\widetilde{T}_i = T_i + N_i$ (without modulo), then we have:
    \begin{align*}
      \left| 2^{k-t}\widetilde{T}_i-U_i \right| &= \left| 2^{k-t}(T_i + N_i)-U_i \right| 
                                                     = \left| 2^{k-t}N_i - U_i[0,k-t] \right| \\
                                                   &\leq \left| 2^{k-t}N_i\rvert + \lvert U_i[0,k-t] \right| 
                                                   % &\leq 2^{k-\ell} + 2^{k-\ell} - 1\\
      \leq 2^{k-t +1 }.
    \end{align*}

  \item Otherwise, there are two possible ``wraparound'' cases:
    \begin{itemize}
    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^t-1$. 
      In this case, we have $U_i = U_i[0,k-t]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
      \[
        \left| 2^{k-t}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-t}(2^t -1) -(U_i + 2^k) \right| =
        \left| 2^{k-t} + U_i[0,k-t]\right| \leq 2^{k-t + 1}. 
      \]

    \item Or $T_i = 2^{t}-1$ and $N_i = +1$, which leads to
      $\widetilde{T}_i = 0$.  This implies that
      $U_i= 2^k-2^{k-t} + U_i[0,k-t]$; we set $U'_i = U_i - 2^k$ (again,
      with this modification $U'$ stays in the lattice), and we find:
      \[
        \left| 2^{k-t}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
        = \left| 2^k-2^{k-t} + U_i[0,k-t] - 2^k\right| \leq 2^{k-t + 1}.
      \]
    \end{itemize}
  \end{itemize}
  In the end, we have
  $\left\| U' -2^{k-t} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-t}$,
  and $U' \equiv U \mod 2^k $.
\end{proof}

@@

Let $\CVP(\mathcal{L}, x)$ denote the vector of $\mathcal{L}$ closest to the
input vector $x$. Using the same notations as above, we want to know if
$\CVP\left(\mathcal{L}, 2^{k-t}\widetilde{T}\right)$ is indeed $U'$. This
will necessarily be the case when $\|2^{k-t}\widetilde{T} - U'\|$ is smaller
than the length of the shortest non-zero vector of $\mathcal{L}$ --- this quantity,
the \emph{first minimum} of the lattice, is denoted by $\lambda_1(\mathcal{L})$.
%What can we say about $|CVP(L,2^{k-\ell}\widetilde{T})-U'|$ ?
By the triangular inequality, we have:
\begin{align*}
|\CVP(\mathcal{L}, 2^{k-t}\widetilde{T}) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-t}\widetilde{T} \right) - 2^{k-t}\widetilde{T}\right | + \left|2^{k-t}\widetilde{T} - U'\right|.
\end{align*}
But, as the vector $U'$ belongs to the lattice, by definition of the closest
vector and by lemma~\ref{thm:close}:
% $|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leq
% |2^{k-\ell}\widetilde{T}-U'| $, hence
\[
  |\CVP(\mathcal{L}, 2^{k-t}\widetilde{T}) - U'| \leq 2 \left|2^{k-t} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-t}.
\]

If we can prove that the right side of this inequality is smaller than the first
minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
$\CVP(\mathcal{L}, 2^{k-t}\widetilde{T})$ indeed reveals $U_0 \bmod 2^k$.

In section~\ref{sec:big_cvp}, we will be facing the problem of reconstructing a
geometric sequence modulo $2^{128}$ given arbitrarily many (noisy versions of
the) most-significant $6$ bits of successive elements of the sequence. Therefore
we have $k=128$ and $t=6$, and we wish to determine the required number of
samples, \textit{i.e.} the value of $n$. This means finding the values of $n$
such that $\sqrt{n} 2^{124} \leq \lambda_1\left(\mathcal{L}\right)$.

Starting from $n=\lceil 122/6\rceil$, we computed the length of the
shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
until the condition holds. The \textsc{Shortest Vector Problem} (SVP) is another
well-known lattice NP-hard problem; we used the (almost) off-the-shelf
\texttt{G6K} library~\cite{AlbrechtDHKPS19}, which gave results very quickly by
sieving. \texttt{fplll}~\cite{fplll} was too slow above dimension 50, in the
default settings.

After this computation, we found that the minimal possible $n$ is 63: with
$n=63$, the shortest vector of $\mathcal{L}$ has length greater than
$2^{127.02}$, which is high enough. This vector can be obtained by bootstrapping
the geometric sequence with
\[
  U_0 = 12144252875850345479015002205241987363
\]
then reducing the terms modulo $2^{128}$ in zero-centered representation
(subtracting $2^{128}$ to $U_i$ if $U_i > 2^{127}$). It follows that when
$n\geq 63, k=128$ and $t=6$, any CVP oracle will return a vector congruent to
the original $U$ when given $\widetilde{T}$.

@@

Denote again by $\mathcal{L}$ the $n$-dimensional lattice spanned by the rows of
$G_{n,64}$, and let $H$ denote the LLL-reduction of $G_{n,64}$. The same lattice
is also spanned by the rows of $H$. For instance, with $n=3$:
\[
  H = \left(\begin{array}{rrr}
                -1241281756092 & 3827459685972 & -728312298332 \\
                -5001120657083 & -2117155768935 & 5479732607037 \\
                8655886039732 & 3303731088004 & 6319848582548
              \end{array}\right)
\]
Set $S = 2^{k-t} \widetilde{T}H^{-1}$; as $2^{k-t}\widetilde{T}$ does not belong
\textit{a priori} to the lattice, $S$ does not have to be an integer
vector. Let then $R$ denote the rounding of $S$, \textit{i.e.}
$R_i = \left\lfloor S_i \right\rceil$. Then $R H$ is an element of the
lattice. Under the right conditions, it will be the vector of the lattice
closest to $2^{k-t} \widetilde{T}$. Indeed:
\begin{align*}
\lVert U' - R H \rVert &= \left\| U' - \left(R - 2^{k-t}\widetilde{T}H^{-1} + 2^{k-t}\widetilde{T}H^{-1}\right)H \right\| \\
&= \left\| U' - 2^{k-t}\widetilde{T} - \left(R - 2^{k-t} \widetilde{T}H^{-1}\right )H \right\| \\
&\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\| R - 2^{k-t}\widetilde{T}H^{-1} \right\| \times \vvvert H \vvvert.
\end{align*}
By definition, $R$ is the closest integer vector to
$2^{k-t}\widetilde{T}H^{-1}$. But as $U'$ is an element of the lattice,
$U'H^{-1}$ is an integer vector. Thus $R-2^{k-t}\widetilde{T}H^{-1}$ is
shorter than $U'H^{-1}-2^{k-t}\widetilde{T}H^{-1}$. Hence :
\begin{align*}
  \left\| U' - RH \right\| &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|R - 2^{k-t}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
                        &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|U'H^{-1}-2^{k-t}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
                        &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|U' - 2^{k-t}\widetilde{T}\right\| \times \left\vvvert H^{-1} \right\vvvert  \times \vvvert H \vvvert \\
                        & 	\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| \times \left(1 + \vvvert H^{-1} \vvvert  \times \vvvert H \vvvert \right).
\end{align*}

Note that $\vvvert H^{-1} \vvvert \times \vvvert H \vvvert$ is the condition
number of the matrix $H$. Lattice reduction has the side effect of reducing the
condition number, therefore it makes sense to use an LLL-reduced basis. If we
can prove that the right side of the inequality is smaller than the first
minimum of the lattice, then we would have proved that $RH$ is indeed the
closest vector we were searching for. Because we have fixed $k = 64$, by
lemma~\ref{thm:close} we have
$\left\| 2^{k-t}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-t}-1
$. So, if we fix $n$ we can search for the minimum number $t$ of known
most-significant bits such that:
\[
  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-t} \leq \lambda_1(\mathcal{L})
\]

\begin{table}[htbp!]
  \centering
  \begin{tabular}{|c|Sc|Sc|c|}
	\hline
	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $t$ \\
	\hline
	3 & 2.87 & $4.09e^{12} \simeq 2^{41.9}$ & 26 \\
	4 & 2.06 & $2.44e^{14} \simeq 2^{47.8}$ & 20 \\
	5 & 3.77 & $1.72e^{15}\simeq 2^{50.6}$ & 18 \\
	6 & 2.69 & $1.03e^{16}\simeq 2^{53.2}$ & 15 \\
	\hline
  \end{tabular}
  \caption{minimal $t$ needed for a given $n$}
  \label{tab:babai}
\end{table}

When $t$ is greater than the values given in table~\ref{tab:babai}, then
Babai's rounding technique will always return the closest vector, and allow us
to reconstruct a truncated geometric series.

@@

Adapting the previous reasoning enables an efficient state reconstruction
algorithm for the \texttt{lehmer64()} generator shown in
Fig.~\ref{fig:prngs}. When clocked, it outputs the top 64 bits of a geometric
sequence ($k=128$ and $t=64$). Three successive outputs are sufficient to
reconstruct the internal state using Babai's rounding technique. This yields the
following reconstruction algorithm:
\begin{minted}[fontsize=\scriptsize]{python}
def reconstruct(X):
    """
    Produce the internal state of the generator given three consecutive outputs of lehmer64().
    16 multiplications, 1 division, 11 additions and 3 roundings only.
    """
    a = 0xda942042e4dd58b5
    r = round(2.64929081169728e-7 * X[0] + 3.51729342107376e-7 * X[1] + 3.89110109147656e-8 * X[2])
    s = round(3.12752538137199e-7 * X[0] - 1.00664345453760e-7 * X[1] - 2.16685184476959e-7 * X[2])
    t = round(3.54263598631140e-8 * X[0] - 2.05535734808162e-7 * X[1] + 2.73269247090513e-7 * X[2])
    u = r * 1556524 + s * 2249380 + t * 1561981
    v = r * 8429177212358078682 + s * 4111469003616164778 + t * 3562247178301810180
    state = (a*u + v) % (2**128)
    return state
\end{minted}

@@

This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
dozen arithmetic operations; it always outputs the correct value of $S_0$, and
may output a few other ones (they can be easily discarded by checking $X_2$). An
improved ``guess-and-determine'' state reconstruction algorithm is possible,
which essentially amounts to expose a truncated version of the underlying linear
congruential generator, and attack it using the tools exposed in
section~\ref{sec:geometric}. This is possible by combining the following
ingredients:
\begin{itemize}
\item The underlying linear congruential generator uses a power-of-two modulus,
  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
  the $\ell$ low-order bits of $S_i$. More precisely, we have:
  \begin{equation}\label{eq:lcg}
    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
  \end{equation}
  Therefore, guessing the least-significant bits of $S_0$ yields a ``long-term
  advantage'' that holds for all subsequent states.

@@

We consider the sequence of internal states
$S = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
least-significant bits of $S_0$, therefore let us assume that their value is
known and denote it by $w$. We define $S' = \LCG_{128}(S_0 - w, 0)$ and
$K = \LCG_{128}(w, c)$ --- this is known. By~\eqref{eq:lcg-additive}, we have
$S' = S - K$. The point is that the elements of $S'$ follow a geometric
progression of common ratio $a$; in addition, the $\ell$ least significant bits
of each components are equal to zero. It follows that $S'[\ell:64+\ell]$ also
follows a geometric progression of common ratio $a$, this time modulo
$2^{64}$. The crux of the reconstruction algorithm is to find
$S'[\ell:64+\ell]$.

We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
unknown vector $B$ of \emph{borrows}, whose components are either 0 or 1, such
that $T' = S[58:64+\ell] \boxminus K[58:64+\ell] \boxminus B$. Because the
borrows are unknown, we in fact compute
$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, and clearly
$\widetilde{T'} = T' \boxplus B$. We are thus in the context of the problem
discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
sequence given $t = 6+\ell$ (noisy) most-significant bits. The ``noise'' is the
unknown vector $B$ of borrows. %Notice that the noise is asymmetric.
